---
title: "Reproducible Research Project 2"
author: "Dhruba"
date: "September 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Reproducible Research Project 2

## Health and Economic Impact of Weather Events in the US

### Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

### This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## 1.  Synopsis

###  The dataset used in this analysis is the U.S. NOAA storm database which tracks characteristics of major storms and severe weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries and property damage. The questions we will be answering are which events are the most harmful with respect to the population health and economy. The analysis will start with data transformation and end with quantitative analysis of the imppact via plot illustration. Results reveal "tornadoes" followed by "excessive heat" are the most dangerous event wrt to population health. On the economic front "flash floods" and "thunderstorm" wreck havoc. Crops are majorly damaged by "drought", "flood", "hails".

### Mention worthy, the data set has skewed data in favour of tornadoes. It is only since 1996 that all events has been recorded. That is why the results are skewed in favour( or against) tornadoes.

##2. Data Processing

```{r}
### Loading the Dataset into R environment
library(data.table)
storm <- fread('StormData.csv', header = T, sep = ',')
```

### Events are aggregated in the next section as required for further analysis.

```{r}
# number of unique event types
length(unique(storm$EVTYPE))

# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)

# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))

# update the data frame
storm$EVTYPE <- event_types
```

## 3. Results 
## Dangerous Events wrt Population Health

### The number of casualties, both injuries and fatalities are aggregated event wise.
#### Plyr package is installed from CRAN for a good visual experience.

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

### Table Representation of Fatalities
```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

### Table Representation of Injuries
```{r}
injury_events[, c("EVTYPE", "injuries")]
```

## Economic damages

### The available property damage and crop damage reportings/estimates were used to estimate economic damages.

```{r}
exp_transform <- function(e) {
  # h -> hundred, k -> thousand, m -> million, b -> billion
  if (e %in% c('h', 'H'))
    return(2)
  else if (e %in% c('k', 'K'))
    return(3)
  else if (e %in% c('m', 'M'))
    return(6)
  else if (e %in% c('b', 'B'))
    return(9)
  else if (!is.na(as.numeric(e))) # if a digit
    return(as.numeric(e))
  else if (e %in% c('', '-', '?', '+'))
    return(0)
  else {
    stop("Invalid exponent value.")
  }
}
```

### Compiling the results to form the datasets which will be used for the visualization of the results. 
```{r}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```


```{r}
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```


```{r}
### Table Representation of Property Damage
prop_dmg_events[, c("EVTYPE", "prop_dmg")]

### Table Representation of Crop Damage
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

##4. Visulazing the results

## cowplot is used to align the graphs side by side to show the effect of diasters wrt to injuries and fatalities side by side.

## Impact on Population Health

### Top dangerous weather event types.

```{r}
library(ggplot2)

# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
  geom_bar(stat="identity") +
  coord_flip() +
  ylab("Total number of fatalities") +
  xlab("Event type") +
  theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
  geom_bar(stat="identity") +
  coord_flip() + 
  ylab("Total number of injuries") +
  xlab("Event type") +
  theme(legend.position="none")


library(cowplot)
plot_grid(p1, p2, align='v')
```

### Tornadoes, Excessive Heat and Flash Floods are the most dangerous events leading to the maximum number of fatalities. On the injury front tornado still is the most dominant force, followed by tstm wind and flood. Overall tornadoes are a major force of havoc.

## Economic Impact

### Plots shows the most damage-prone disasters from 1950s

```{r}
library(ggplot2)

# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("Event type") +
  ylab("Property damage in dollars (log-scale)") +
  theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
  geom_bar(stat="identity") +
  coord_flip() + 
  xlab("Event type") +
  ylab("Crop damage in dollars") + 
  theme(legend.position="none")

p1
```

### Flood , hurricane and tornado are the major sources of devastation on the front of property damage.

```{r}
p2
```

### The graph shows the damage on crops on the major disasters. As can be guessed by intuition, drought appears to be the major source of destruction on crops followed by flood, where as river flood and ice storm are in major contestants of the 3rd place.