---
title: "Reproducible Research Project 2"
author: "Dhruba"
date: "September 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Reproducible Research Project 2

## Health and Economic Impact of Weather Events in the US

### Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

### This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Synopsis

### The analysis on the storm event database revealed that tornadoes are the most dangerous weather event to the population health. The second most dangerous event type is the excessive heat. The economic impact of weather events was also analyzed. Flash floods and thunderstorm winds caused billions of dollars in property damages between 1950 and 2011. The largest crop damage caused by drought, followed by flood and hails.



## Loading File

```{r}
library(data.table)
storm <- fread('StormData.csv', header = T, sep = ',')
```


```{r}
# number of unique event types
length(unique(storm$EVTYPE))
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
# update the data frame
storm$EVTYPE <- event_types
```

## Dangerous Events with respect to Population Health

### To find the event types that are most harmful to population health, the number of casualties are aggregated by the event type.

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```


```{r eval=FALSE}
fatal_events[, c("EVTYPE", "fatalities")]
```


```{r}
injury_events[, c("EVTYPE", "injuries")]
```

## Economic Effects of Weather Events

### To analyze the impact of weather events on the economy, available property damage and crop damage reportings/estimates were used.

```{r}
exp_transform <- function(e) {
  # h -> hundred, k -> thousand, m -> million, b -> billion
  if (e %in% c('h', 'H'))
    return(2)
  else if (e %in% c('k', 'K'))
    return(3)
  else if (e %in% c('m', 'M'))
    return(6)
  else if (e %in% c('b', 'B'))
    return(9)
  else if (!is.na(as.numeric(e))) # if a digit
    return(as.numeric(e))
  else if (e %in% c('', '-', '?', '+'))
    return(0)
  else {
    stop("Invalid exponent value.")
  }
}
```


```{r}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```


```{r}
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```


```{r}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

## RESULTS

## Health impact of weather events

### The following plot shows top dangerous weather event types.

```{r}
library(ggplot2)

# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
  geom_bar(stat="identity") +
  coord_flip() +
  ylab("Total number of fatalities") +
  xlab("Event type") +
  theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
  geom_bar(stat="identity") +
  coord_flip() + 
  ylab("Total number of injuries") +
  xlab("Event type") +
  theme(legend.position="none")


library(cowplot)
plot_grid(p1, p2, align='h')
```

### Tornadoes cause most number of deaths and injuries among all event types. There are more than 5,000 deaths and more than 10,000 injuries in the last 60 years in US, due to tornadoes. The other event types that are most dangerous with respect to population health are excessive heat and flash floods.

## Economic Impact of weather Events
### The following plot shows the most severe weather event types with respect to economic cost that they have costed since 1950s

```{r}
library(ggplot2)

# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("Event type") +
  ylab("Property damage in dollars (log-scale)") +
  theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
  geom_bar(stat="identity") +
  coord_flip() + 
  xlab("Event type") +
  ylab("Crop damage in dollars") + 
  theme(legend.position="none")

library(cowplot)
plot_grid(p1, p2, align='h')
```
